# Summary of tools

## R libraries

The following libraries are required for this to run:

```{r, echo=T}
library(matrixStats)
library(yaml)
library(ggplot2)
library(ggrepel)
library(reshape2)
library(DT)
library(kableExtra)
library(GenomicFeatures)
library(RColorBrewer)
library(gplots)
library(GSEABase)
library(copynumber)
library(readxl)
library(survival)
library(forestplot)
# library(YAPSA)
library(EGSEA)
library(org.Hs.eg.db)
library(ggbeeswarm)
library(dplyr)
library(BSgenome.Hsapiens.UCSC.hg19)
hs <- org.Hs.eg.db
source("scripts/GenomeLocationPlot.R")
source("scripts/CalcCNVs.R")
source("scripts/ContingencyTablePlot.R")
source("scripts/Process_for_MutPanning.R")
AllDat <- read_yaml("input_data.yml")
```

## Analysis Tools

Summary:

Process | Tools
--------|------------
Alignment | bwa-mem and Novosort
QC | Picard metrics
Tumor SNV | Strelka, Vardict, Mutect2
Germline SNV | Haplotypecaller, Vardict
SNV Annotations | VEP
CNV | cnvkit
Ploidy & purity | Gentien

## Annotation data

### Target Bed 

Note a targeted panel was used for this analysis, which is loaded into the variable `TargetBed`

```{r}
TargetBed <- read.delim(AllDat$Annot$Bed, header = F, sep = "\t")
TargetBedGR <- GRanges(seqnames = TargetBed$V1, ranges = IRanges(start = TargetBed$V2, end = TargetBed$V3), Gene = TargetBed$V4)
TargetBedGR$width <- width(TargetBedGR)
genesB <- unique(TargetBed$V4)
genesB2 <- genesB[-grep("_[0-9]+", genesB)]
```

There are `r length(unique(TargetBed$V4))` targets and `r nrow(TargetBed)` intervals in this panel featuring `r length(genesB2)`.

Below is a summary table of chromosomal locations and genes in close proximity (using a sliding window of 10MB):

```{r}
Windows <- GenomeLocationSeg(AllDat$Annot$Bed, AllDat$Annot$HgChr, windsize = 1E7)
WindowList <- Windows[[1]]
BedList <- Windows[[2]]
WindowM <- melt(WindowList)
colnames(WindowM) <- c("Chr", "Start", "End", "Width", "Strand", "NGenes", "Genes", "Site", "Col")
DT::datatable(WindowM[which(WindowM$Genes != ""), c("Chr", "Start", "End", "Genes")], rownames = F, class = "cell-border stripe", extensions = "Buttons", options = list(dom = "Bfrtip", buttons = c("csv", "excel")))

WindowPlot(WindowM, "NGenes")
```

### Cancer Gene Census

The cancer gene census was downloaded from [cosmic](https://cancer.sanger.ac.uk/cosmic/download). This can be loaded to filter out oesophagael specific genes. There are two tables loaded: `CmcOesoph` which is an Oseophageal specific Cosmic List, and `CmcHallmark` which is  List of Cosmic Hallmark genes

```{r cload}
CosmicList <- read.csv(AllDat$Annot$CancerCensus)
CmcOesoph <- CosmicList[grep("oesoph", CosmicList$Tumour.Types.Somatic.), ]
CmcHallmark <- CosmicList[unique(c(
     grep("oesoph", CosmicList$Tumour.Types.Somatic.),
     which(CosmicList$Hallmark == "Yes")
)), ]
```

Pan-TCGA pathways, extracted from [pathway mapper](https://github.com/iVis-at-Bilkent/pathway-mapper), and used to visualise abberant pathways in cbioportal is saved in the variable `TCGApathways`

```{r}
TCGApathways <- read.csv(AllDat$Annot$TCGApathway, stringsAsFactors = F)
tmpnam <- colnames(TCGApathways)
TCGApathways <- lapply(1:ncol(TCGApathways), function(x) setdiff(unique(TCGApathways[, x]), ""))
names(TCGApathways) <- tmpnam
```

<!--
### GeneSets and Ontologies

Currently not being loaded... using Hallmark pathway analysis?

```{r, eval=T}
PathInc2 <- getGmt(
     con = "annot/c2.cp.v7.4.symbols.gmt", geneIdType = SymbolIdentifier(),
     collectionType = BroadCollection(category = "c2")
)

PathInH <- getGmt(
     con = "annot/h.all.v7.4.symbols.gmt", geneIdType = SymbolIdentifier(),
     collectionType = BroadCollection(category = "h")
)
```
-->

# Data

All external dependencies and data files are summarised in the following yaml file:

```{r}
kable(unlist(AllDat), col.names = "Summary of inputs")
```

## Patient Samples

Below is a list of the samples used in this analysis and their clinical information saved in the variable `SampleIDs`.

Need to factor specific inputs for later survival analysis

```{r datasumm}
SampleIDs <- read.delim(AllDat$SampleList, sep = "\t")
SampleIDs$Gender.Male.1..F.0 <- factor(SampleIDs$Gender.Male.1..F.0)
# SampleIDs$Cancer.Location.By.Pathologist=factor(SampleIDs$Cancer.Location.By.Pathologist)
SampleIDs$Grade <- factor(SampleIDs$Grade, levels = c("Poorly diff", "Medium", "Highly diff", "Mucinous"))
SampleIDs$Stage <- factor(SampleIDs$Stage)
SampleIDs$pT <- factor(SampleIDs$pT)
SampleIDs$pN <- factor(SampleIDs$pN)
SampleIDs$pM <- factor(SampleIDs$pM)
SampleIDs$Margin.status...R0.R2 <- factor(SampleIDs$Margin.status...R0.R2)
SampleIDs$Stage <- factor(substr(SampleIDs$AJCC_8, 1, 1))
SampleIDs$Alive[which(SampleIDs$Alive == T)] <- "Alive"
SampleIDs$Alive[which(SampleIDs$Alive == 0)] <- "Deceased"
SampleIDs$Location <- factor(SampleIDs$Cancer.Location.By.Pathologist) # SampleIDs$Cancer.Location.By.Pathologist
SampleIDs$Outcome[which(SampleIDs$Outcome == "")] <- NA
SampleIDs$Age.at.op <- as.numeric(SampleIDs$Age.at.op)
SampleIDs$OutCol <- factor(SampleIDs$Outcome)
levels(SampleIDs$OutCol) <- c("grey75", "darkcyan", "red", "forestgreen", "darkorange")
```

There are `r nrow(SampleIDs)` samples in this cohort.

The location of these tumors are either the cardia (near stomach, Siewert II) or distal esophagus (Siewert I), locations where EAC are more common, and these have all been classified as adenocarcinomas by pathology (EAC).

## Somatic Single Nucleotide Variants

Somatic variants were either called with:

* strelka where matched normal was present to create `StTab1`
* mutect2 `M2Tab1`
* vardict `varGTab1`

```{r variantOverview}
########### LOAD STRELKA DATA

StFiles <- dir(AllDat$strelka)

StrelkaAll <- lapply(StFiles, function(x) read.delim(paste0(AllDat$strelka, "/", x), sep = "\t"))

namIdx <- sapply(strsplit(StFiles, "_|.maf"), function(x) x[2])

names(StrelkaAll) <- SampleIDs$SAMPLE_ID[match(namIdx, SampleIDs$SAMPLE_ID)]
meltStrelkaAll <- melt(StrelkaAll, measure.vars = "DISTANCE")

cNames <- c("CHROM", "POS", "REF", "ALT", "STRAND", "Hugo_Symbol", "cadd_phred", "HGVSp_Short", "Consequence", "Tumor_Sample_Barcode", "gn_af", "nAlt.TUMOR", "nTot.TUMOR", "VAF.TUMOR", "mgrb_af")
Tab1 <- meltStrelkaAll[, match(cNames, colnames(meltStrelkaAll))]
Tab1$gn_af <- round(Tab1$gn_af, 2)

StTab1 <- Tab1

########### LOAD MUTECT DATA

StFiles <- dir(AllDat$Mutect2)

mutect2All <- lapply(StFiles, function(x) read.delim(paste0(AllDat$Mutect2, "/", x), sep = "\t"))

namIdx <- sapply(strsplit(StFiles, "_|.maf"), function(x) x[2])

names(mutect2All) <- SampleIDs$SAMPLE_ID[match(namIdx, SampleIDs$SAMPLE_ID)]
meltmutect2All <- melt(mutect2All, measure.vars = "DISTANCE")

cNames <- c("CHROM", "POS", "REF", "ALT", "Hugo_Symbol", "cadd_phred", "HGVSp_Short", "Consequence", "gn_af", "X1k_af", "mgrb_af", "max_af", "PolyPhen", "STRAND", "nAlt.TUMOR", "nTot.TUMOR", "VAF.TUMOR", "nAlt.NORMAL", "nTot.NORMAL", "VAF.NORMAL", "Tumor_Sample_Barcode")
Tab1 <- meltmutect2All[, match(cNames, colnames(meltmutect2All))]
Tab1$gn_af <- round(Tab1$gn_af, 2)

M2Tab1 <- Tab1

########### LOAD VARDICT DATA

VDFiles <- dir(AllDat$vardict)

varAll <- lapply(VDFiles, function(x) read.delim(paste0(AllDat$vardict, "/", x), sep = "\t"))

namIdx <- sapply(strsplit(VDFiles, "_|.maf"), function(x) x[2])

names(varAll) <- SampleIDs$SAMPLE_ID[match(namIdx, SampleIDs$SAMPLE_ID)]
meltvdAll <- melt(varAll, measure.vars = "DISTANCE")


cNames <- c("CHROM", "POS", "REF", "ALT", "Hugo_Symbol", "cadd_phred", "HGVSp_Short", "Consequence", "gn_af", "X1k_af", "mgrb_af", "max_af", "PolyPhen", "STRAND", "nAlt.TUMOR", "nTot.TUMOR", "VAF.TUMOR", "nAlt.NORMAL", "nTot.NORMAL", "VAF.NORMAL", "Tumor_Sample_Barcode")
GTab1 <- meltvdAll[, match(cNames, colnames(meltvdAll))]
GTab1$gn_af <- round(GTab1$gn_af, 2)

# Filter out variants which have a gnomad frequency of greater than 75% in gnomad and mgrb
tx2 <- which(GTab1$gn_af > 0.75 | GTab1$mgrb_af > 0.75)

varGTab1 <- GTab1[-tx2, ]
```


## Germline SN Variants

Germline variants were either called with HaplotypeCaller (`HCTab1`) or Vardict (`GTab1`). Both were annotated with VEP

```{r}
# Haplotype caller here
VDFiles <- dir(AllDat$HCcaller)

hcAll <- lapply(VDFiles, function(x) read.delim(paste0(AllDat$HCcaller, "/", x), sep = "\t"))

namIdx <- sapply(strsplit(VDFiles, "_|.maf"), function(x) x[7])

names(varAll) <- namIdx ## SampleIDs$SAMPLE_ID[match(namIdx, SampleIDs$SAMPLE_ID)]
melthcAll <- melt(hcAll, measure.vars = "DISTANCE")


cNames <- c("CHROM", "POS", "ID", "REF", "ALT", "Hugo_Symbol", "cadd_phred", "HGVSp_Short", "Consequence", "L1", "gn_af", "X1k_af", "mgrb_af", "max_af", "PolyPhen", "cadd_phred.1", "DOMAINS", "CLIN_SIG", "STRAND", "nAlt.NORMAL", "nTot.NORMAL", "VAF.NORMAL", "Tumor_Sample_Barcode")
HCTab1 <- melthcAll[, match(cNames, colnames(melthcAll))]
HCTab1$gn_af <- round(HCTab1$gn_af, 2)
# HCTab1$END=HCTab1$POS+mVals-1
# HCTab1$Variant_Type=ifelse(mVals==1, "SNP",ifelse(a1>a2, "DEL", "INS"))

# find the deleterious consequences?
# # enter the types of variants we want to report
# CodingVar=c("stop", "miss", "frameshift", "prime")
# delCons=unlist(sapply(CodingVar, function(x) grep(x, GTab1$Consequence)))

# Load vardict results here

VDFiles <- dir(AllDat$vardictGermline)

varAll <- lapply(VDFiles, function(x) read.delim(paste0(AllDat$vardictGermline, "/", x), sep = "\t"))

namIdx <- sapply(strsplit(VDFiles, "_|.maf"), function(x) x[2])

names(varAll) <- namIdx ## SampleIDs$SAMPLE_ID[match(namIdx, SampleIDs$SAMPLE_ID)]
meltvarAll <- melt(varAll, measure.vars = "DISTANCE")

## save this data table\
# a1=nchar(meltvarAll$REF)
# a2=nchar(meltvarAll$ALT)
# mVals=rowMaxs(cbind(a1, a2))

cNames <- c("CHROM", "POS", "ID", "REF", "ALT", "Hugo_Symbol", "cadd_phred", "HGVSp_Short", "Consequence", "L1", "gn_af", "X1k_af", "mgrb_af", "max_af", "PolyPhen", "cadd_phred", "CLIN_SIG", "STRAND", "nAlt.NORMAL", "nTot.NORMAL", "VAF.NORMAL", "Tumor_Sample_Barcode")


GTab1 <- meltvarAll[, match(cNames, colnames(meltvarAll))]
GTab1$gn_af <- round(GTab1$gn_af, 2)
```


## CNVs

* Note CNVkit calls as follows: https://cnvkit.readthedocs.io/en/stable/pipeline.html#call

* Normals are combined to create a panel of normals for comparisons, but calls were conducted on a single-sample level, reported as `CNVlog2`
* LOH was inferred by gentian and loaded as `lohInfo`

```{r CNVkit}
CNVfiles <- dir(AllDat$CNVkit)

CNVlog2 <- lapply(CNVfiles, function(x) read.delim(paste0(AllDat$CNVkit, "/", x), sep = "\t"))
namIdx <- sapply(strsplit(CNVfiles, "_|-gainloss.tsv"), function(x) x[1])
# t2=match(gsub("-", ".", namIdx), SampleIDs$SAMPLE_ID)
names(CNVlog2) <- SampleIDs$SAMPLE_ID[match(namIdx, sapply(strsplit(SampleIDs$SAMPLE_ID, "\\."), function(x) x[1]))]

# Add CNV calls at a gene level
pathview <- "data/cnv_genes_gentian/"
filesIn <- dir(pathview)
dnaSegGentian <- lapply(paste(pathview, filesIn, sep = ""), function(x) read.csv(x))
names(dnaSegGentian) <- sapply(dnaSegGentian, function(x) x$Sampleid[1])
```


```{r loh}
loh <- dir(AllDat$loh)

lohInfo <- lapply(loh, function(x) read.csv(paste0(AllDat$loh, "/", x)))
names(lohInfo) <- sapply(strsplit(loh, "_"), function(x) x[1])
```


## Estimates of purity

Purity was estimated using PureCN. Below is a plot of the inferred ploidy compared to inferred purity across the cohort in the variable `pinfo`

```{r loadpurity}
pfiles <- dir(AllDat$purity)
pinfo <- sapply(pfiles, function(x) read.csv(paste0(AllDat$purity, "/", x)))
colnames(pinfo) <- SampleIDs$SAMPLE_ID[match(unlist(strsplit(colnames(pinfo), ".csv")), sapply(strsplit(SampleIDs$SAMPLE_ID, "\\."), function(x) x[1]))]

plot(unlist(pinfo["Purity", ]), unlist(pinfo["Ploidy", ]),
     ylab = "inferred ploidy", xlab = "inferred purity"
)
```