# Evaluate Consensus calls

## Germline Variants

Two variant callers are used: HaplotypeCaller (HC, left) and Vardict (VD, right).

Before and filtering, we compare:

* **Variant Allele Frequency (VAF)** - There should be an approximate bimodal distribution with centred around 0.5 (heterozygous) or 1 (homozygous). (Note: many homozygous variants are known SNPs, following filtering, most variants are heterozygous)
* **Variant Allele Frequency (VAF)** - the number of supporting alternate reads at the site 


```{r processGermlineVars}
## do a quick assessment of HC variants and Vardict variants:
par(mfrow = c(2, 2))
hist(HCTab1$VAF.NORMAL, main = "VAF HC")
hist(GTab1$VAF.NORMAL, main = "VAF Var")
hist(HCTab1$nAlt.NORMAL, main = "nAlt HC")
hist(GTab1$nAlt.NORMAL, main = "nAlt Var")
```

Also compare the variant frequencies in gnomad and MGRB. Note that these range from 0 to 1 since all variants are called irrespective of whether it is a common variant or not.

```{r}
par(mfrow = c(2, 2))
hist(HCTab1$gn_af, main = "gnomad HC")
hist(GTab1$gn_af, main = "gnomad Var")
hist(HCTab1$mgrb_af, main = "mgrb HC")
hist(GTab1$mgrb_af, main = "mgrb Var")
```

**We apply the following thresholds:**

* MGRB OR GNOMAD allelic frequency of 1\% or less (consider NA values as 0)
* At least 5 reads supporting each variant
* Variant in a coding region (this includes 5'UTR and 3'UTR regions and splice variants)
* Present in both Haplotype caller and Vardict
* Exclude the 6 samples which failed sequencing QC

This data is saved in the final variable `FinalGermlineList`

```{r}
FREQ_CUTOFF <- 0.01
CODING_TERMS <- c(
  "stop", "miss", "frameshift", "prime",
  "splice[a|d|r]", "protein", "coding_sequence"
)

# Find all unique variants
HCind <- variant_key(HCTab1)
VDind <- variant_key(GTab1)

## Merge all samples together and filter out gnomad variants of a particular frequency:
FinalGermlineList <- HCTab1[which(HCind %in% VDind), ]

# Make a note of common genes - these could potentially be germline
GermSumm <- paste(FinalGermlineList$CHROM, FinalGermlineList$POS)
Xa <- table(GermSumm)
PotentialGermlineGene <- names(Xa)[which(Xa > 5)]

## Remove failed samples
FinalGermlineList <- FinalGermlineList[
  !(FinalGermlineList$Tumor_Sample_Barcode %in% FAILED_SAMPLES),
]

## Frequency filter
FinalGermlineList <- af_filter(FinalGermlineList, FREQ_CUTOFF)

## Annotate the variants which are functional/coding
FinalGermlineList$Functional <- grepl(
  paste(CODING_TERMS, collapse = "|"),
  FinalGermlineList$Consequence
)
```

We can visualise the VAF, nALT, gnomad and mgrb population freq distrbutions in this merged dataset compared to the orignal data frames:

```{r}
par(mfrow = c(2, 2))
boxplot(HCTab1$VAF.NORMAL, GTab1$VAF.NORMAL, FinalGermlineList$VAF.NORMAL, main = "VAF", names = c("HC1", "VD1", "Combined"))
boxplot(HCTab1$nAlt.NORMAL, GTab1$nAlt.NORMAL, FinalGermlineList$VAF.NORMAL, main = "nALT", names = c("HC1", "VD1", "Combined"), log = "y")
boxplot(HCTab1$gn_af, GTab1$gn_af, FinalGermlineList$gn_af, main = "gn_AF", names = c("HC1", "VD1", "Combined"))
boxplot(HCTab1$mgrb_af, GTab1$mgrb_af, FinalGermlineList$mgrb_af, main = "MGRB_AF", names = c("HC1", "VD1", "Combined"))
```

Finally, consistency between the two callers:

* red: number of variants pre-filtering
* blue : variants remaining after gnomad filtering
* purple : final number of variants taking into account coding only variants
* grey: final list

```{r GermlineOutputBars}
samples_norm <- sort(na.omit(SampleIDs$NORMAL_ID))

HCNvar <- table(HCTab1$Tumor_Sample_Barcode)
VDNvar <- table(GTab1$Tumor_Sample_Barcode)

overlap_counts <- count_overlap(
  HCind, HCTab1$Tumor_Sample_Barcode,
  VDind, GTab1$Tumor_Sample_Barcode,
  samples_norm
)

TableInts <- data.frame(
  Sample = samples_norm,
  HCOrig = as.numeric(as.character(HCNvar[samples_norm])),
  VardictOrig = as.numeric(as.character(VDNvar[samples_norm])),
  Overlap = overlap_counts,
  stringsAsFactors = FALSE
)

TableInts[is.na(TableInts)] <- 0

final_counts <- table(FinalGermlineList$Tumor_Sample_Barcode)

TableInts$Final <- as.numeric(as.character(final_counts[TableInts$Sample]))
TableInts$Final[is.na(TableInts$Final)] <- 0

# VDNvar=table(FinalGermlineList$Tumor_Sample_Barcode)
# midx=match(names(VDNvar), TableInts$HCOrig.Var1)
# TableInts$Final=0
# TableInts$Final[midx]=VDNvar

DT::datatable(TableInts)

barplot(
  t(TableInts[, c("HCOrig", "VardictOrig", "Overlap", "Final")] + 1),
  beside = TRUE,
  las    = 2,
  col    = c("red", "blue", "purple", "grey"),
  main   = "Number of germline variants identified using HC",
  log    = "y"
)
```

```{r}
FinalGermline <- split(FinalGermlineList, FinalGermlineList$Tumor_Sample_Barcode)

GRangeGerm <- lapply(FinalGermline, function(x) {
  GRanges(
    seqnames = x$CHROM,
    ranges = IRanges(x$POS, x$POS),
    loh = NA, TumourCN = NA
  )
})

names(GRangeGerm) <- sub("\\..*", "", names(GRangeGerm))

## reorder the lohInfo into a GRanges object
GRangeLOH <- lapply(lohInfo, function(x) {
  GRanges(
    seqnames = x$chr,
    ranges   = IRanges(start = x$start, end = x$end),
    loh      = x$type,
    C        = x$C
  )
})

for (i in seq_along(GRangeGerm)) {
  germ_name <- names(GRangeGerm)[i]
  hits <- findOverlaps(GRangeGerm[[i]], GRangeLOH[[germ_name]])
  GRangeGerm[[i]]$loh[queryHits(hits)] <- GRangeLOH[[germ_name]]$loh[subjectHits(hits)]
  GRangeGerm[[i]]$TumourCN[queryHits(hits)] <- GRangeLOH[[germ_name]]$C[subjectHits(hits)]
}

FinalGermline <- Map(
  cbind,
  FinalGermline,
  lapply(GRangeGerm, mcols)
)

FinalGermlineList <- do.call(rbind, FinalGermline)
```

Below we can review these variants and print to file:

```{r}
DT::datatable(FinalGermlineList,
  rownames = F, class = "cell-border stripe",
  extensions = "Buttons", options = list(dom = "Bfrtip", buttons = c("csv", "excel"), scrollX = T)
)
write.table(FinalGermlineList, file = "output/Final_germline_mutations_test_new.maf", sep = "\t", row.names = F)
```


## Evaluate consensus tumor

Perform the same analysis for the tumor: We want to firstly see the number of variants idnetified with each caller:

```{r SomaticVarsOutput}
# Create the unique variant keys
Sind <- variant_key(StTab1)
M2ind <- variant_key(M2Tab1)
VTind <- variant_key(varGTab1)

# get the unique tumor ids
samples_tum <- sort(na.omit(SampleIDs$SAMPLE_ID))

## Compute Pairwise overlaps
intValSM <- count_overlap(
  Sind, StTab1$Tumor_Sample_Barcode,
  M2ind, M2Tab1$Tumor_Sample_Barcode,
  samples_tum
)

intValSV <- count_overlap(
  Sind, StTab1$Tumor_Sample_Barcode,
  VTind, varGTab1$Tumor_Sample_Barcode,
  samples_tum
)

intValMV <- count_overlap(
  M2ind, M2Tab1$Tumor_Sample_Barcode,
  VTind, varGTab1$Tumor_Sample_Barcode,
  samples_tum
)

# calculate the unique number per sample
Ssum <- table(StTab1$Tumor_Sample_Barcode)
M2sum <- table(M2Tab1$Tumor_Sample_Barcode)
VTsum <- table(varGTab1$Tumor_Sample_Barcode)

# calculate the # muts per sample as well as the filtered number
TableInts <- data.frame(
  sample = samples_tum, St = Ssum[samples_tum],
  M2 = M2sum[samples_tum], Vardict = VTsum[samples_tum],
  SM = intValSM[samples_tum], SV = intValSV[samples_tum],
  MV = intValMV[samples_tum]
)
TableInts <- TableInts[order(TableInts$sample), ]
```

Overview of number of variants:

* red: number of Strelka variants
* blue: Mutect2 variants
* purple: vardict variants

```{r}
barplot(t(TableInts[, c(3, 5, 7)]),
  beside = T, las = 2, col = c("red", "blue", "purple"), names.arg = TableInts$sample, cex.names = 0.75, log = "y",
  ylab = "number of somatic variants identified"
)
```

Overview of the intersection between different methods:

* red: number of Strelka & M2variants
* blue: Strelka & Vardict
* purple: M2 & vardict

```{r}
# intersection between different methods
barplot(t(TableInts[, c(8:10)]),
  beside = T, las = 2, col = c("red", "blue", "purple"), names.arg = TableInts$sample, cex.names = 0.75,
  ylab = "intersection between different methods"
)
```
For mutpanning, keep all the variants

* Variants called by two of strelka, mutect2 or vardict

Below, calculate which variants are called by multiple tools. Then take the mutect variants which have at least 1 other caller, and all vardict variants which are also called by strelka. We save this table as `FinalSomaticList`.

```{r}
M2Tab1$Mutect <- 1
M2Tab1$St <- NA
M2Tab1$VD <- NA
M2Tab1$St <- as.integer(M2ind %in% Sind)
M2Tab1$VD <- as.integer(M2ind %in% VTind)
M2Tab1$RSums <- rowSums(M2Tab1[, c("Mutect", "St", "VD")], na.rm = T)

varGTab1$Mutect <- NA
varGTab1$St <- NA
varGTab1$VD <- 1
varGTab1$St <- as.integer(VTind %in% Sind)
varGTab1$Mutect <- as.integer(VTind %in% M2ind)
varGTab1$RSums <- rowSums(varGTab1[, c("Mutect", "St", "VD")], na.rm = T)

FinalSomaticList <- rbind(
  M2Tab1[M2Tab1$RSums > 1, ],
  varGTab1[varGTab1$RSums == 2 & varGTab1$Mutect == 0, ]
)
```

The final list of somatic variants:
* filtering out common gnomad or mgrb variants of 0.01 
* Annotate whether a variant is **potential germline** - the tumour has a high allele fraction (50\%+) AND the variant is detected in the GermlineList. OR VAF which are too high in non-matched normal samples (These would be likely 70\% or lower purity, High VAF of 80+ are possibly germline variants)
* Annotate whether a variant is potential **Normal expansion** - the variant is detected at a very low frequency in the corresponding normal sample, but high in the tumour 


```{r}
FinalSomaticList <- af_filter(FinalSomaticList, FREQ_CUTOFF)

## Create a column for "potential germline", or normal expansion
isPotentialGermline <- (paste(FinalSomaticList$CHROM, FinalSomaticList$POS) %in% PotentialGermlineGene &
  FinalSomaticList$VAF.TUMOR > 0.5) | (FinalSomaticList$VAF.TUMOR > 0.75)

isNormalExpansion <- FinalSomaticList$VAF.NORMAL < 0.1 &
  FinalSomaticList$nAlt.NORMAL > 3 &
  FinalSomaticList$VAF.TUMOR > 0.3

# Assign to columns
FinalSomaticList$PotentialGermline <- as.integer(isPotentialGermline)
FinalSomaticList$NormalExpansion <- as.integer(isNormalExpansion)
```



```{r}
FinalSomatic <- split(FinalSomaticList, FinalSomaticList$Tumor_Sample_Barcode)

GRangeSomatic <- lapply(FinalSomatic, function(x) {
  GRanges(
    seqnames = x$CHROM,
    ranges = IRanges(x$POS, x$POS),
    loh = NA, TumourCN = NA, gene_loh = NA, gene_C = NA
  )
})

names(GRangeSomatic) <- sub("\\..*", "", names(GRangeSomatic))

## reorder the lohInfo into a GRanges object
GRangeLOH <- lapply(lohInfo, function(x) {
  GRanges(
    seqnames = x$chr,
    ranges   = IRanges(start = x$start, end = x$end),
    loh      = x$type,
    C        = x$C
  )
})

## reorder dnaSegGentian into a GRanges object
GRangeGentian <- lapply(dnaSegGentian, function(x) {
  GRanges(
    seqnames = x$chr,
    ranges   = IRanges(start = x$start, end = x$end),
    loh      = x$loh,
    C        = x$C
  )
})

for (i in seq_along(GRangeSomatic)) {
  germ_name <- names(GRangeSomatic)[i]
  hits <- findOverlaps(GRangeSomatic[[i]], GRangeLOH[[germ_name]])
  GRangeSomatic[[i]]$loh[queryHits(hits)] <- GRangeLOH[[germ_name]]$loh[subjectHits(hits)]
  GRangeSomatic[[i]]$TumourCN[queryHits(hits)] <- GRangeLOH[[germ_name]]$C[subjectHits(hits)]

  if (germ_name %in% names(GRangeGentian)) {
    hits_gentian <- findOverlaps(GRangeSomatic[[i]], GRangeGentian[[germ_name]])
    GRangeSomatic[[i]]$gene_loh[queryHits(hits_gentian)] <- GRangeGentian[[germ_name]]$loh[subjectHits(hits_gentian)]
    GRangeSomatic[[i]]$gene_C[queryHits(hits_gentian)] <- GRangeGentian[[germ_name]]$C[subjectHits(hits_gentian)]
  }
}


FinalSomatic <- Map(
  cbind,
  FinalSomatic,
  lapply(GRangeSomatic, mcols)
)

FinalSomaticList <- do.call(rbind, FinalSomatic)
```

Below we can review these variants and print to file:

```{r}
DT::datatable(FinalSomaticList,
  rownames = F, class = "cell-border stripe",
  extensions = "Buttons", options = list(dom = "Bfrtip", buttons = c("csv", "excel"), scrollX = T)
)
write.table(FinalSomaticList, file = "output/Final_somatic_mutations_test_new.maf", sep = "\t", row.names = F)
```


## Summary of number of variants per sample

```{r summary-variants}
SomVal <- summary(factor(FinalSomaticList$Tumor_Sample_Barcode, levels = SampleIDs$SAMPLE_ID))
GermVal <- summary(factor(FinalGermlineList$Tumor_Sample_Barcode, levels = SampleIDs$NORMAL_ID))

names(SomVal) <- sapply(strsplit(names(SomVal), "\\."), function(x) x[1])
names(GermVal) <- sapply(strsplit(names(GermVal), "\\."), function(x) x[1])

dfSumm <- data.frame(
  Sample = names(SomVal), Somatic = SomVal, Germ = GermVal[match(names(SomVal), names(GermVal))], Alive_edited = S2$Alive_edited[match(names(SomVal), S2$Name)],
  FollowUp = S2m$Follow.up.Survival..months.[match(names(SomVal), S2m$Name)],
  Alive = S2m$Alive[match(names(SomVal), S2m$Name)]
)

mdfSumm <- melt(dfSumm, measure.vars = c("Somatic", "Germ"))
mdfSumm$Sample <- factor(mdfSumm$Sample, levels = S2$Name)

nxs <- median(SomVal)
nxg <- median(GermVal)

highMB <- SampleIDs$SAMPLE_ID[match(names(SomVal)[which(SomVal > 100)], SampleIDs$Name)]
```

Notes: 

* median number of somatic variants `r nxs`
* median number of germline variants `r nxg`
* 5 outlier samples with high number of mutations: These include `r names(SomVal)[which(SomVal>100)]`

```{r summary-barplot}
ggplot(data = mdfSumm, aes(x = Sample, y = value, col = value)) +
  geom_bar(stat = "identity") +
  facet_grid(variable ~ Alive, space = "free_x", scale = "free") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  ggtitle("Survival data: stage colored") +
  ylab("Number of variants") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  geom_hline(yintercept = c(nxs))
```