# Annotation of Driver Mutations

A number of different sources were used to identify driver mutations:

1. [MutPanning](https://www.genepattern.org/modules/docs/MutPanning#gsc.tab=0) : prediction of driver mutation based on frequency of a variant in a non-synonymous region compared to background noise for a given gene
2. [OncodriveFML](http://bbglab.irbbarcelona.org/oncodrivefml/home): prediction of driver mutation based on frequency across coding and non-coding regions
3. [CGI annotations](https://www.cancergenomeinterpreter.org/) : to add information of predicted oncogenicity, role (activating vs lof) and potential drugs which can contribute to response
4. [Cosmic Based annotations for treatment](https://cancer.sanger.ac.uk/cosmic/drug_resistance) and CMC consensus Mutations
5. [oncokb](https://github.com/oncokb/oncokb-annotator): MSK resource listing known driver mutations and drug indications in disease specific fashion

## Mutpanning

The following code implements processing such that the output `.maf` file can be loaded into mutpanning for detection of driver events. 

Note that the input requires:
- a liftover of consequence annotations (see function `map_consequence_annotations`)
- an end position and annotation as SNP, DEL, INS
- onco tree metadata annotation 

```{r mutpanning-save}
# save for mut panning
# Process somatic variants
SaveOutput <- process_variants_mutpanning(FinalSomaticList, include_hgvsp = TRUE)
SaveOutput <- format_maf_output(SaveOutput, include_hgvsp = TRUE)

# Process germline variants
SaveOutputG <- process_variants_mutpanning(FinalGermlineList, include_hgvsp = FALSE)
SaveOutputG <- format_maf_output(SaveOutputG, include_hgvsp = FALSE)

# Write somatic mutations to file
write.table( SaveOutput, file = paste0("output/Oesoph_somatic_mutations_for_mutpannning_", Sys.Date(), ".maf"),
  sep = "\t",row.names = FALSE,quote = FALSE
)

# Write somatic mutations to file
write.table( SaveOutputG, file = paste0("output/Oesoph_germline_mutations_for_mutpannning_", Sys.Date(), ".maf"),
  sep = "\t",row.names = FALSE,quote = FALSE
)

# Create and write sample metadata
sampleData <- create_sample_metadata(SampleIDs)
write.table(sampleData,file = paste0("output/Oesoph_somatic_sample_for_mutpannning_", Sys.Date(), ".txt"),
  sep = "\t",row.names = FALSE, quote = FALSE
)

```

Below are the top driver candidates from analysis using mutpanning:

```{r}
mutPanningRes=read.delim(AllDat$DriverMuts$MutPanning, header=T, sep="\t")
head(mutPanningRes)
```

QQ plot for mutpanning, indicating the top significant predicted driver genes

```{r}
Pvals=-log10(mutPanningRes$Significance+0.01)
Pvals[which(Pvals<0)]=0
GeneNames=mutPanningRes$Name
GeneNames[which(Pvals<1)]=NA
dfB=data.frame(name=GeneNames, x=Pvals)

ggplot(data = dfB, aes(sample = x)) + geom_qq() + geom_text_repel(label=dfB$name[order(dfB$x)], stat="qq") + stat_qq_line()+xlab("theoretical quantiles")+ylab("sample quantiles")
```

We can add this information into the FinalSomaticList object. Any gene which is in the MutPanning list can be annotated as a driver event

```{r}
FinalSomaticList$MutPanning=NA
FinalSomaticList$MutPanning[which(FinalSomaticList$Hugo_Symbol%in%GeneNames)]="driver"
```

Print out a summary of the number of driver genes:

```{r}
table(FinalSomaticList$MutPanning, useNA="always")
```

## OncodriveFML

Oncodrive detects potential coding and promoter variants. 

```{r}
onocoFML1=read.delim(AllDat$DriverMuts$CodingFML)
onocoFML2=read.delim(AllDat$DriverMuts$PromoterFML)
tempD=rbind(cbind(onocoFML1, type="Coding"), cbind(onocoFML2, type="Promoter"))
test1=read.delim(AllDat$Annot$Ens2Hugo, sep="\t")
M1=test1$Approved.symbol[match(tempD$GENE_ID, test1$Ensembl.gene.ID)]
tempD$Approved.symbol=M1

DT::datatable(tempD)
```

A similar analysis creating qq-plots for oncodrive FML promoter and coding regions indicating significant genes:

![Promoter rankings](`r AllDat$DriverMuts$PromoterFMLPlot`)

![Coding rankings](`r AllDat$DriverMuts$CodingFMLPlot`)

<!--
DO THIS LATER

We can Also plot pvalues between mutpanning and oncodriveFML to check for consistency in estimating driver events:

```{r}
mx2=match(tempD$SYMBOL, mutPanningRes$Name)
mx3=match(tempD$Approved.symbol, mutPanningRes$Name)
mx2[which(is.na(mx2))]=mx3[which(is.na(mx2))]

mutList=mutPanningRes[mx2, ]

tempD$Symbol2=tempD$Approved.symbol
tempD$Symbol2[which(tempD$P_VALUE>0.01 & mutList$Significance>0.05)]=NA

plot(-log10(mutList$Significance), -log10(tempD$P_VALUE), col=factor(tempD$type), xlab="mutPanning", ylab="oncodriveFML", pch=20)
text(-log10(mutList$Significance)+0.05, -log10(tempD$P_VALUE)+0.05, tempD$Symbol2,cex=0.5)
```

Report a summary of which of these genes are potential drivers

```{r}
head(tempD[which(!is.na(tempD$Symbol2)), ])
## get the list of genes which are in either mutpanning or oncodriveFML
sigGenes=na.omit(tempD$Symbol2)
```

-->

Select the genes which have a Pvalue less than 0.1 and list them as potential Oncodrive driver events

```{r}
OncodriveSig=tempD[which(tempD$P_VALUE<0.1), c("Approved.symbol", "type")]
FinalSomaticList$Oncodrive=NA
FinalSomaticList$Oncodrive[which(FinalSomaticList$Hugo_Symbol%in%OncodriveSig$Approved.symbol[OncodriveSig$type=="Coding"])]="Coding"
FinalSomaticList$Oncodrive[which(FinalSomaticList$Hugo_Symbol%in%OncodriveSig$Approved.symbol[OncodriveSig$type=="Promoter"])]="Promoter"
```

Print a table of the number of Coding and Promoter genes:

```{r}
table(FinalSomaticList$Oncodrive, useNA="always")
```

## CGI annotations

- [CGI annotations](https://www.cancergenomeinterpreter.org/) : to add information of predicted oncogenicity, role (activating vs lof) and potential drugs which can contribute to response. Note this is on a mutation specific level, rather than gene level using mutpanning and OncodriveFML

Below is the mutation information:

```{r}
SNVCGI=read.delim(AllDat$CGI$snv, sep="\t")
DT::datatable(SNVCGI)
```

And the drug information

```{r}
Drug=read.delim(AllDat$CGI$drug, sep="\t")
Drug$Gene2=sapply(strsplit(Drug$SAMPLE_ALTERATION, " "), function(x) x[1])
DT::datatable(Drug)
```

Append this information into the FinalSomaticList on a specific variant level:

```{r}
# Create lookup key and match CGI data
lookup_key <- paste(FinalSomaticList$CHROM, FinalSomaticList$POS)
cgi_key <- paste(SNVCGI$CHR, SNVCGI$POS_HG19)
match_idx <- match(lookup_key, cgi_key)

# Map CGI annotations
FinalSomaticList$CGI_oncogenic_class <- SNVCGI$ONCOGENIC_CLASSIFICATION[match_idx]
FinalSomaticList$CGI_role <- SNVCGI$CONSENSUS_ROLE[match_idx]

# Identify variants to profile
has_role_and_protein_affecting <- which(
  FinalSomaticList$CGI_role != "" & 
  FinalSomaticList$CGI_oncogenic_class != "non-protein affecting"
)

is_driver_or_predicted <- which(
  FinalSomaticList$CGI_oncogenic_class %in% c("known driver", "predicted")
)

profile_indices <- unique(c(has_role_and_protein_affecting, is_driver_or_predicted))

# Initialize drug response columns
FinalSomaticList$CGIResistance <- NA
FinalSomaticList$CGIResponse <- NA

# Process drug associations
for (i in profile_indices) {
  # Filter drugs for this variant
  variant_drugs <- Drug[
    Drug$SAMPLE == FinalSomaticList$Tumor_Sample_Barcode[i] & 
    Drug$Gene2 == FinalSomaticList$Hugo_Symbol[i],
  ]
  
  # Get complete matches for cancer type
  complete_matches <- variant_drugs[
    variant_drugs$CTYPE_MATCH == "YES" & 
    variant_drugs$ALTERATION_MATCH == "complete",
  ]
  
  if (nrow(complete_matches) > 0) {
    # Extract responsive and resistant drugs
    responsive_drugs <- unique(complete_matches$DRUGS[complete_matches$RESPONSE == "Responsive"])
    resistant_drugs <- unique(complete_matches$DRUGS[complete_matches$RESPONSE == "Resistant"])
    
    FinalSomaticList$CGIResponse[i] <- paste(responsive_drugs, collapse = " ")
    FinalSomaticList$CGIResistance[i] <- paste(resistant_drugs, collapse = " ")
  }
}
```

<!--
```{r, eval=F}
## The old code to remove
Lx1=match(paste(FinalSomaticList$CHROM, FinalSomaticList$POS), paste(SNVCGI$CHR, SNVCGI$POS_HG19))

FinalSomaticList$CGI_oncogenic_class=SNVCGI$ONCOGENIC_CLASSIFICATION[Lx1]
FinalSomaticList$CGI_role=SNVCGI$CONSENSUS_ROLE[Lx1]

Tx2=which(FinalSomaticList$CGI_role!="" & FinalSomaticList$CGI_oncogenic_class!="non-protein affecting")
Tx3=which(FinalSomaticList$CGI_oncogenic_class%in% c("known driver", "predicted"))
ProfileThese=unique(c(Tx2,Tx3))

FinalSomaticList$CGIResistance=NA
FinalSomaticList$CGIResponse=NA

for (i in 1:length(ProfileThese)){
  temp2=Drug[which(Drug$SAMPLE==FinalSomaticList$Tumor_Sample_Barcode[ProfileThese[i]] & 
                     Drug$Gene2==FinalSomaticList$Hugo_Symbol[ProfileThese[i]]), ]
  la=which(temp2$CTYPE_MATCH=="YES" & temp2$ALTERATION_MATCH=="complete")
  if (length(la)>0) {
    temp2b=temp2[la, ]
    resp=paste(unique(temp2b$DRUGS[which(temp2b$RESPONSE=="Responsive")]), collapse=" ")
    resist=paste(unique(temp2b$DRUGS[which(temp2b$RESPONSE=="Resistant")]), collapse=" ")
    FinalSomaticList$CGIResponse[ProfileThese[i]]=resp
    FinalSomaticList$CGIResistance[ProfileThese[i]]=resist
  }
}
```
-->

From this data base the predicted number of activating, loss of function and ambiguous variants are indicated below:

```{r}
table(FinalSomaticList$CGI_role)
```

We can also look at the output response information:

```{r}
table(FinalSomaticList$CGIResponse)
```

And resistance information

```{r}
table(FinalSomaticList$CGIResponse)
```

## Cosmic

There are two pieces of information to add here. 

- [Cosmic Based annotations for treatment](https://cancer.sanger.ac.uk/cosmic/drug_resistance): information on drug indications against specific mutations
- [CMC mutational Census](https://cancer.sanger.ac.uk/cmc/home): consensus on driver mutations based on prevalence across 1500 cancer types

```{r}
# Read and prepare mutation data
ResMut <- read.delim(AllDat$Annot$CosmicMut)
ResMut$Symbol <- sub("_.*", "", ResMut$Gene.Name)  # Simpler than strsplit/sapply

# Create variant keys for matching
ResMut$variant_key <- paste(ResMut$Symbol, ResMut$AA.Mutation)
FinalSomaticList$variant_key <- paste(FinalSomaticList$Hugo_Symbol, FinalSomaticList$HGVSp_Short)
FinalGermlineList$variant_key <- paste(FinalGermlineList$Hugo_Symbol, FinalGermlineList$HGVSp_Short)

# Match somatic variants and aggregate drug names
somatic_matches <- match(ResMut$variant_key, FinalSomaticList$variant_key)
matched_muts <- ResMut[!is.na(somatic_matches), c("variant_key", "Drug.Name")]

# Aggregate drugs by variant (simpler than by/stack)
drug_lookup <- aggregate(Drug.Name ~ variant_key, 
                         data = matched_muts, 
                         FUN = paste, 
                         collapse = " ")

# Add drugs to FinalSomaticList
FinalSomaticList$CosmicDrugs <- drug_lookup$Drug.Name[match(FinalSomaticList$variant_key, drug_lookup$variant_key)]

# Match germline variants
germline_matches <- match(ResMut$variant_key, FinalGermlineList$variant_key)
nxa <- which(!is.na(germline_matches))

```


<!--
```{r, eval=F}
ResMut=read.delim(AllDat$Annot$CosmicMut)
ResMut$Symbol=sapply(strsplit(ResMut$Gene.Name, "_"), function(x) x[1])

midx1=match(paste(ResMut$Symbol, ResMut$AA.Mutation), paste(FinalSomaticList$Hugo_Symbol, FinalSomaticList$HGVSp_Short))
nxa=which(!is.na(midx1))
Lx2=FinalSomaticList[unique(midx1[nxa]), ]

mList=ResMut[nxa ,c("Symbol", "AA.Mutation", "Drug.Name")]
mList=unique(mList)

pasteVariants=stack(by(mList$Drug.Name,paste(mList$Symbol, mList$AA.Mutation),  paste, collapse=" "))

FinalSomaticList$CosmicDrugs=NA
FinalSomaticList$CosmicDrugs[unique(midx1[nxa])]=pasteVariants$values[match(paste(Lx2$Hugo_Symbol, Lx2$HGVSp_Short), pasteVariants$ind)]

midx2=match(paste(ResMut$Symbol, ResMut$AA.Mutation), paste(FinalGermlineList$Hugo_Symbol, FinalGermlineList$HGVSp_Short))
nxa=which(!is.na(midx2))
```
-->

```{r cmc-load, cache=T}
#load("annot/CancerMutCensus_LVL1-3.RData")
load(AllDat$Annot$CancerMutCensusCache)
```

```{r}
CMCv92short$hg37Start=sapply(strsplit( CMCv92short$Mutation.genome.position.GRCh37, ":"), function(x) x[2])
CMCv92short$hg37Start=sapply(strsplit( CMCv92short$hg37Start, "-"), function(x) x[1])

midx=match(paste(FinalSomaticList$Hugo_Symbol, FinalSomaticList$HGVSp_Short), paste(CMCv92short$GENE_NAME, CMCv92short$Mutation.AA))
CMCv92shortf=CMCv92short[na.omit(midx), ]

FinalSomaticList$CMC=NA
FinalSomaticList$CMC[which(!is.na(midx))]=CMCv92shortf$MUTATION_SIGNIFICANCE_TIER
FinalSomaticList$TSG_OCG=NA
FinalSomaticList$TSG_OCG[which(!is.na(midx))]=ifelse(CMCv92shortf$ONC_TSG%in%c("oncogene", "oncogene, fusion"), "onc", "tsg")
```

The number of variants which land in a CMC tier (1 being highest) is indicated below:

```{r}
table(FinalSomaticList$CMC)
```

Similarly, we can report whether they are known tumour suppressors or oncogenes:

```{r}
table(FinalSomaticList$TSG_OCG)
```

```{r}
DT::datatable(FinalSomaticList[1:100, ], rownames=F, class='cell-border stripe', extensions="Buttons", options=list(dom="Bfrtip", buttons=c('csv', 'excel')))
```

## Oncokb

- [oncokb](https://github.com/oncokb/oncokb-annotator): MSK resource for known oncogenes and drug indications agains them

To Do: Arrange genes in order of frequency in good vs bad response?

```{r addoncokb}
OncokbDFs=read.delim(AllDat$OncokbAnnotations, sep="\t")
##AllDat$OncokbAnnotations
OncokbSNV=read.delim(AllDat$DriverMuts$SNVFML, sep="\t")

# create variant key match
FinalSomaticList$variant_key <- variant_key(FinalSomaticList)
OncokbSNV$variant_key <- variant_key(OncokbSNV) #paste(OncokbSNV$Chromosome, OncokbSNV$Start_Position, OncokbSNV$Tumor_Sample_Barcode, sep=":")
match_idx <- match(FinalSomaticList$variant_key, OncokbSNV$variant_key)

# Initialize columns
FinalSomaticList$ONCOKB_class  <- NA
FinalSomaticList$ONCOKB_effect <- NA
FinalSomaticList$OncokbDrugs   <- NA
FinalSomaticList$Drugs         <- NA

# Collapse LEVEL_ columns per variant
 LevelDF=OncokbSNV[ , grep("LEVEL_", colnames(OncokbSNV))]
 LevelDFCOll=apply(LevelDF, 1 , paste , collapse = "-" )
 LevelDFCOll=gsub("NA-", "", LevelDFCOll)
 LevelDFCOll=gsub("[-]+NA", "", LevelDFCOll)
 names(LevelDFCOll)=OncokbSNV$variant_key
 
 
 # Fill in OncoKB annotations for matching variants
FinalSomaticList$ONCOKB_class[!is.na(match_idx)]  <- OncokbSNV$ONCOGENIC[match_idx[!is.na(match_idx)]]
FinalSomaticList$ONCOKB_class[FinalSomaticList$ONCOKB_class == "Unknown"]   <- NA

FinalSomaticList$ONCOKB_effect[!is.na(match_idx)] <- OncokbSNV$MUTATION_EFFECT[match_idx[!is.na(match_idx)]]
FinalSomaticList$ONCOKB_effect[FinalSomaticList$ONCOKB_effect == "Unknown"] <- NA

FinalSomaticList$OncokbDrugs[!is.na(match_idx)] <- LevelDFCOll[match_idx[!is.na(match_idx)]]
# Combine with CosmicDrugs if available
ndx <- which(!is.na(FinalSomaticList$OncokbDrugs))
FinalSomaticList$Drugs[ndx] <- paste(FinalSomaticList$CosmicDrugs[ndx], FinalSomaticList$OncokbDrugs[ndx])
```

Below, print out the oncogenic and mutation effect information

```{r}
table(OncokbSNV$ONCOGENIC)
table(OncokbSNV$MUTATION_EFFECT)
```

