# Summary of somatic variants

## Most common variants

Below, we collapse the variants such that the most common in this cohort are reported. 

* dark blue: Cosmic oesophagael specific
* light blue: Cosmic cancer genes

```{r most-common-variants, fig.cap="Frequency of coding mutations in assessed genes. Skyblue: cancer gene, Darkblue: oesoph specific"}
CodingVar <- "splice_(donor|acceptor|region)|stop|missense|frameshift|prime"

Nmut <- sapply(Maf2GRangesFiltgn, function(x) x[grepl(CodingVar, x$Consequence, ignore.case = TRUE)])
# Nmut=sapply(Maf2GRangesFiltgn, function(x) x[(which(!is.na(x$HGVSp)& substr(x$HGVSp, nchar(x$HGVSp), nchar(x$HGVSp))!="="))])

C2 <- do.call("c", Nmut)
df.Coding <- data.frame(CHR = seqnames(C2), POS = start(ranges(C2)), REF = C2$REF, ALT = C2$ALT, Gene = C2$Gene, HGVSp = C2$HGVSp, Consequence = C2$Consequence, Samp = C2$Samp)

Output1 <- acast(df.Coding[, c("Gene", "Samp")], Gene ~ Samp)
Nx <- names(FinalSomatic)[(!names(FinalSomatic) %in% colnames(Output1))]
Mat2 <- matrix(0, ncol = length(Nx), nrow = nrow(Output1))
colnames(Mat2) <- Nx
Output1 <- cbind(Output1, Mat2)

# calculate the frequency of variant, ignoring multiple occurrences in the same sample
Output1Freq <- rowSums(sign(Output1)) / ncol(Output1)
x1 <- rowSums(sign(Output1))
x1b <- rowSums(Output1[, -which(colnames(Output1) %in% highMB)])
x2 <- sort(x1, decreasing = T)

ColId <- ifelse(names(x2) %in% DriverEvents$Gene, "skyblue", "grey")
ColId[which(names(x2) %in% DriverEvents$Gene[which(!DriverEvents$Cons %in% c("Unknown", "Unknown-novel"))])] <- "slateblue1"

barplot(x2[1:25] / 64, las = 2, ylab = "Frequency of coding mutation", col = ColId, ylim = c(0, 1))
```

Note that TP53 is the most ccommon mutation across the cohort

## Mutational frequencies compared to other cohorts (Fig2A)

Mutational frequency in our cohort was comparable to:

* [MSK2020](https://www.cbioportal.org/study/summary?id=egc_mskcc_2020)
* [TCGA ESCA legacy](https://www.cbioportal.org/study/summary?id=esca_tcga)
* [TCGA 2017 publication](https://www.cbioportal.org/study/summary?id=stes_tcga_pub)
* [DFCI2013](https://www.cbioportal.org/study/summary?id=esca_broad) Nature Genetics 2013


```{r Fig2A-comapre-frequencies, fig.cap="Comparison of frequency of mutations in current cohort with other published methods"}
# Load comparison cohort data files
comp_cohort_files <- AllDat$CompCohort
comp_cohort_names <- unlist(strsplit(basename(unlist(comp_cohort_files)), "_Mutated_Genes.txt"))

# Read and store comparison data in a named list
comp_data_list <- lapply(comp_cohort_files, function(f) read.delim(f, sep = "\t"))
names(comp_data_list) <- comp_cohort_names

# Initialize frequency table with current cohort data
df1 <- data.frame(Genes = names(Output1Freq), LundEAC = Output1Freq)

# Merge frequencies from other cohorts
for (cohort_name in names(comp_data_list)) {
  cohort_df <- comp_data_list[[cohort_name]]

  # Clean percentage strings and extract frequencies
  cohort_freqs <- cohort_df %>%
    dplyr::select(Gene, Freq) %>%
    dplyr::mutate(!!cohort_name := as.numeric(sub("%", "", Freq)) / 100) %>%
    dplyr::select(-Freq)

  df1 <- merge(df1, cohort_freqs, by.x = "Genes", by.y = "Gene", all.x = TRUE)
}

# Order by current cohort frequency
df1b <- df1[order(df1$LundEAC, decreasing = TRUE), ]

# Frequency check for variants outside top 25 (e.g., > 15% in other cohorts)
# Using arr.ind=T to find indices where frequency > 0.15
high_freq_others <- which(df1b[-c(1:25), -1] > 0.15, arr.ind = TRUE)

# Do a proportion test to check if there is a difference in frequency.
# cohort_sizes: LundEAC (64), MSK2020 (286), TCGA (79), DFCI (151)
cohort_sizes <- c(64, 286, 79, 151)

calculate_prop_test <- function(row) {
  # Extract frequencies and convert to counts
  freqs <- as.numeric(row[2:5])
  freqs[is.na(freqs)] <- 0
  counts <- freqs * cohort_sizes

  # Run proportion test
  tryCatch(
    {
      prop.test(counts, cohort_sizes)$p.value
    },
    error = function(e) NA
  )
}

pval_list <- apply(df1b, 1, calculate_prop_test)
names(pval_list) <- df1b$Genes

# Identify genes with significant differences after FDR correction
sig_genes_fdr <- names(which(p.adjust(pval_list, method = "fdr") < 0.05))

# Prepare top 25 genes for plotting
df_plot <- df1b[1:25, ]
df_plot$Genes_fmt <- as.character(df_plot$Genes)
df_plot$Genes_fmt[df_plot$Genes %in% sig_genes_fdr] <- paste0(df_plot$Genes[df_plot$Genes %in% sig_genes_fdr], "*")

melted_freqs <- melt(df_plot[, !colnames(df_plot) %in% "Genes"], id.vars = "Genes_fmt")
melted_freqs$Genes_fmt <- factor(melted_freqs$Genes_fmt, levels = df_plot$Genes_fmt)

# Define color coding for gene labels
gene_colors <- ifelse(df_plot$Genes %in% DriverEvents$Gene, "darkblue",
  ifelse(df_plot$Genes %in% CmcOesoph$Gene.Symbol, "blue",
    ifelse(df_plot$Genes %in% CosmicList$Gene.Symbol, "lightblue", "black")
  )
)

ggplot(melted_freqs, aes(x = Genes_fmt, y = value, col = variable)) +
  geom_point() +
  scale_color_manual(values = c("red", "skyblue", "#8c6bb1", "#7fcdbb")) +
  labs(x = "Gene", y = "Frequency", color = "Cohort") +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, color = gene_colors),
    axis.title.x = element_blank()
  )

# Generate and save summary table for export
df_mut_freq <- df1b
df_mut_freq$is_oesophageal <- df_mut_freq$Genes %in% CmcOesoph$Gene.Symbol
df_mut_freq$is_predicted_driver <- df_mut_freq$Genes %in% DriverEvents$Gene
df_mut_freq$p_value <- pval_list[match(df_mut_freq$Genes, names(pval_list))]
df_mut_freq$fdr_p_value <- p.adjust(pval_list, method = "fdr")[match(df_mut_freq$Genes, names(pval_list))]
df_mut_freq$is_significant_fdr <- df_mut_freq$Genes %in% sig_genes_fdr

write.csv(df_mut_freq, file = "output/Fig2A_frequencies_stats.csv", row.names = FALSE)
```

The significant genes have a p-value less than 0.05 based on a proportionality test (indicated with an asterisk * above), and thus appear at a different frequency compared to other public data.


## Association of mutational status with outcome

Create a univariate cox model, looking at each mutation independently but accounting for age, stage, location, sex etc:

* TP53 associated with good outcome
* RYR1, ARID1A and CDKN2A with poorer outcome

```{r Fig3A, message=FALSE, warning=FALSE}
# Select genes with at least 8 mutations for univariate survival analysis
min_mutations <- 8
frequent_genes <- names(x2)[x2 >= min_mutations]

# Prepare data frame for survival analysis
# Combine clinical data (S2) with mutation status for selected genes
mutation_status <- t(sign(Output1[frequent_genes, match(S2$SAMPLE_ID, colnames(Output1)), drop = FALSE]))
survival_data <- cbind(S2, mutation_status)

# Define survival object (using follow-up time and deceased status)
surv_obj <- Surv(survival_data$Follow.up.Survival..months., survival_data$Alive == "Deceased")

# Run Cox proportional hazards model for each gene independently
# Adjusting for clinical covariates: Age, Stage, Grade, Location, Gender, LN.Ratio
run_cox_model <- function(gene_name, data, surv) {
  # Formula: clinical covariates + the gene's mutation status
  formula_str <- paste0("surv ~ log(Age.at.op) + Stage + Grade + Location + Gender.Male.1..F.0 + LN.Ratio + `", gene_name, "`")

  tryCatch(
    {
      model <- coxph(as.formula(formula_str), data = data)
      summary_model <- summary(model)

      # Extract coefficients and confidence intervals for the target gene (last row)
      coef_data <- summary_model$coefficients
      conf_int <- summary_model$conf.int
      last_row <- nrow(coef_data)

      data.frame(
        Gene = gene_name,
        Coeff = coef_data[last_row, "exp(coef)"], # Hazard Ratio
        Lower = conf_int[last_row, "lower .95"],
        Upper = conf_int[last_row, "upper .95"],
        P = coef_data[last_row, "Pr(>|z|)"],
        Psum = summary_model$logtest["pvalue"],
        stringsAsFactors = FALSE
      )
    },
    error = function(e) {
      message(paste("Error modeling gene", gene_name, ":", e$message))
      NULL
    }
  )
}

# Iterate over genes and combine results into a single data frame
cox_results <- lapply(frequent_genes, run_cox_model, data = survival_data, surv = surv_obj) %>%
  dplyr::bind_rows()

# Clean up result types for presentation
cox_results$Coeff <- as.numeric(cox_results$Coeff)
cox_results$Lower <- as.numeric(cox_results$Lower)
cox_results$Upper <- as.numeric(cox_results$Upper)
cox_results$P <- round(as.numeric(cox_results$P), 2)

# Visualize results using a forest plot
forestplot(
  cox_results[, c("Gene", "P")],
  cox_results$Coeff,
  cox_results$Lower,
  cox_results$Upper,
  xlog = TRUE,
  boxsize = 0.1
)
```

```{r gene-lists, include=FALSE}
EAClist <- c("ARID1A", "CDKN2A", "SMAD4", "SMARCA4", "PBRM1", "ERBB2", "SPG20", "TLR4", "ELMO1", "DOCK2")
ESCClist <- c("KDM6A", "KMT2C", "KMT2D", "NFE2L2", "ZNF750", "NOTCH1", "TGFBR2", "EGFR", "PIK3R1", "PTCH1", "KEAP1", "CUL3")
Both <- c("TP53", "PIK3CA")
GAlist <- c("CHFR", "MGMT", "APC", "ITGAV", "RUNX1", "FHIT", "MYC", "VEGFA", "RHOA")
Allg <- c(EAClist, Both, GAlist, ESCClist)

# Define variables for downstream section 08 compatibility
MMRgenes <- c("MSH2", "MSH6", "POLE", "MSH3", "MLH1", "MLH3", "PMS2", "PMS1", "MUTYH")
a1_mmr <- which(df.Coding$Gene %in% MMRgenes)
MMRout <- stack(by(df.Coding$Gene[a1_mmr], df.Coding$Samp[a1_mmr], paste))

HRgenes <- c("BRCA1", "BRCA2", "RAD50", "ATM", "H2AX", "PALB2", "RAD52", "MRE11", "BRIP1", "BARD1", "RAD51D")
a1_hr <- which(df.Coding$Gene %in% HRgenes)
HRout <- stack(by(df.Coding$Gene[a1_hr], df.Coding$Samp[a1_hr], paste, collapse = " "))

ApproxCoding <- sum(TargetBedGR$width) / 1E6
Nmut <- table(factor(C2$Samp, levels = names(FinalSomatic)))
TMB <- Nmut / ApproxCoding

dfMut <- data.frame(Sample = names(TMB), TMB = TMB, Ntot = TotGn, MMRgenes = NA, HRgenes = NA, ER = S2$Alive_edited[match(names(TMB), S2$SAMPLE_ID)])
dfMut$MMRgenes[na.omit(match(MMRout$ind, dfMut$Sample))] <- MMRout$values
dfMut$HRgenes[na.omit(match(HRout$ind, dfMut$Sample))] <- HRout$values
dfMut$isMMRorHR <- ifelse(!is.na(dfMut$MMRgenes) | !is.na(dfMut$HRgenes), "yes", "no")
dfMut$Genes <- paste(dfMut$MMRgenes, dfMut$HRgenes)
dfMut$Genes <- gsub("NA", "", dfMut$Genes)
dfMut$Genes[which(dfMut$Genes == " ")] <- NA
```

## Co-occurrence matrix for genes: 

Determine whethere there is any border-line significant associations which can explain lower than expected number of mutations, or convergence in mutational profile. 

The plot below shows association between PIK3CA and KMT2C; RUNX1 and KMT2D and MGMT and PIK3CA

```{r corr-mut}
# l2=match(test, rownames(Output1))
l2 <- c(match(Allg, rownames(Output1)), match(frequent_genes, rownames(Output1)))
t2 <- Output1[na.omit(unique(l2)), ]
Mx1 <- cor(t(t2))
## calculate a fisher's test for co-occurence:
##
# FT=fisher.test(t(t2))
pvals <- sapply(1:nrow(t2), function(x) sapply(1:nrow(t2), function(y) fisher.test(table(t2[x, ], t2[y, ]))$p.value))
# diag(pvals)=1
pvals2 <- p.adjust(pvals, "fdr")
pvals3 <- matrix(pvals2, nrow = nrow(pvals))
colnames(pvals3) <- colnames(Mx1)
rownames(pvals3) <- rownames(Mx1)

dfnew <- melt(Mx1)
dfm2 <- melt(pvals3)
dfnew$p <- dfm2$value
dfnew$sig <- ifelse(dfnew$p < 0.1, "*", ifelse(dfnew$p < 0.05, "**", ""))

ggplot(data = dfnew, aes(x = Var1, y = Var2, fill = value)) +
  geom_tile() +
  scale_fill_gradient2(low = "#2C7BB6", mid = "white", high = "#D7191C") +
  geom_text(aes(label = sig), color = "black", size = 5) +
  labs(y = NULL, x = NULL, fill = "tau") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = -45, hjust = 0))
```


