# CNV data

## Overall gains and losses at a population level (Fig2B)

CNV regions are reported as a log2 fold change compared to the normal sample. Assuming 100% purity, we could expect a gain to have a FC of 0.58,  duplication 1, het loss of (-1)

Using the coding from cnvkit, the following frequency plot uses the following thresholds (0.3), which detects gains and losses at 50% purity

```{r, eval=F}
* Less than -1.1: Hom loss
* -1.1 to -0.4: Het Loss
* 0.3 to 1 gain
* 1+ amp
```

```{r cnv-pop-freq, fig.cap="Overview of common sites of gains and losses in the cohort"}
filesIn=dir(AllDat$CNVkitGenes)
sampinfo=unlist(strsplit(filesIn, "_genes.csv"))
GeneCNVs=lapply(filesIn, function(x) read.csv(paste0(AllDat$CNVkitGenes,"/", x)))
names(GeneCNVs)=SampleIDs$SAMPLE_ID[match(sampinfo, sapply(strsplit(SampleIDs$SAMPLE_ID, "\\."), function(x) x[1]))]
## how to call CNVs here?

meltCN=melt(CNVlog2, measure.vars=c("seg.mean"))
AllDCN=data.frame(sampleID=meltCN$L1, chrom=meltCN$chrom, arm="p", start.pos=meltCN$loc.start,
                  end.pos=meltCN$loc.end, n.probes=meltCN$num.mark, mean2=meltCN$value, CNval=meltCN$C,
                  mean=ifelse(meltCN$C>2,0.4, ifelse(meltCN$C<2, -0.4, 0)))

## manually 
plotFreq(segments=AllDCN,thres.gain=0.3,thres.loss=-0.3, ylim=c(-100,100), main="oesophagael cohort")

```

Given that our purity values appear to range between 0.2 to 0.6 (see section 2), some of these gains are undetected. Using a threshold of 0.16, will allow us to detect a gain at 25% purity (expect log2 values for gains: 0.17, and loss at -0.19 )

This plot roughly matches the frequency plot in the TCGA cohort [Figure 4](https://www.nature.com/articles/nature20805), where common gained regions include:

* 1q
* chr 2, 7, 8

Losses:
* chr 4-5
* chr 18

## Overall gains and losses in good and poor survivors (Fig2B) - Is this needed?

Repeat the above analysis for the three outcome groups - good, poor and intermediate

```{r}

plotFreq(segments=AllDCN[ which(AllDCN$sampleID%in%SampleIDs$SAMPLE_ID[SampleIDs$Alive_edited=="Alive"]), ],thres.gain=0.15,thres.loss=-0.15, ylim=c(-100,100), main="Alive")

plotFreq(segments=AllDCN[ which(AllDCN$sampleID%in%SampleIDs$SAMPLE_ID[SampleIDs$Alive_edited=="Deceased"]), ],thres.gain=0.15,thres.loss=-0.15, ylim=c(-100,100), main="Deceased")

```

## Thresholds for copy number

Here, using the inferred number of copies after adjusting for purity, we can make a call on the CNVs

Using cutoffs:

* hom loss: 0
* het loss: 1
* normal: between 1-3
* gain: 3-6
* amp: 6+

In this section, also write to file the summary information for input into gistic and CGI

```{r, fig.cap="Summary of the distribution of gains and losses relative to segment mean"}
allchr=c(1:22)
meltCN=melt(GeneCNVs, measure.vars=c("seg.mean"))
AllDCNGene=data.frame(sampleID=meltCN$L1, chrom=meltCN$chr, arm="p", start.pos=meltCN$start,
                  end.pos=meltCN$end, n.probes=meltCN$number.targets, mean=meltCN$value, gene=meltCN$gene.symbol, C=meltCN$C, loh=meltCN$loh)

## rank the Gene order based on chromosomal location
AllDCNGene=AllDCNGene[ order(AllDCNGene$chrom, AllDCNGene$start.pos),]
AllDCNGene$gene=factor(AllDCNGene$gene, levels=unique(AllDCNGene$gene))

AllDCNGeneb=AllDCNGene[which(AllDCNGene$gene%in%TargetBed$V4), ]
AllDCNGeneb$cutoff=cut(AllDCNGeneb$C, c(-1,0.01, 1.01, 3,5.99, 1000), c(-2, -1, 0, 1, 2))
AllDCNGeneb$cutoff=as.numeric(as.character(AllDCNGeneb$cutoff))

## write this to file for mafannotator
CNVtable=acast(AllDCNGeneb[ ,c("sampleID", "gene", "cutoff")], gene~sampleID)
CNVtable2=data.frame(Hugo_Symbol=rownames(CNVtable), Locus_ID=0, Cytoband=NA,CNVtable)
CNVtable2$Cytoband=paste0(AllDCNGeneb$chrom[match(CNVtable2$Hugo_Symbol, AllDCNGeneb$gene)], AllDCNGeneb$arm[match(CNVtable2$Hugo_Symbol, AllDCNGeneb$gene)])
Tx2=grep("_", CNVtable2$Hugo_Symbol)
CNVtable2=CNVtable2[-Tx2, ]

write.table(CNVtable2, file=paste0("output/Oesoph_somatic_CNVgistic_", Sys.Date(), ".txt"), sep="\t", row.names = F, quote = F)

temp2=AllDCNGeneb[ which(AllDCNGeneb$cutoff!=0), ]
temp2$cna=ifelse(temp2$cutoff<0, "del", "amp")
outputCNVlist=unique(temp2[ ,c("gene", "cna")])

write.table(outputCNVlist, file=paste0("output/Oesoph_somatic_CNV_CGI", Sys.Date(), ".txt"), sep="\t", row.names = F, quote = F)

smoothScatter(AllDCNGeneb$mean, log2(AllDCNGeneb$C+0.05), xlab="mean", ylab="log 2 copies", main="distribution of CN calls vs segment FC")
rect(-5, log2(1+0.05), 5,log2(2),  col=rgb(0.2, 0.2, 0.2, alpha=0.2), border = NA)
abline(h=log2(0.05), col="red")
abline(h=log2(1), col="red")
abline(h=log2(6), col="red")
abline(h=log2(3), col="red")
text(-3, 2, "gains")
text(-3, 4, "amp")
text(-3, 1, "diploid range")
text(-3,-4, "homo loss")
text(-3,-1, "het loss")

```


```{r , fig.cap="sites of high frequency aberrations"}
## Below is a summary of the sites with high frequency of gains and losses: The filled color is the segmented mean, and the box outline color is adjusted for ploidy:

Xl=lapply(allchr, function(x) acast(AllDCNGeneb[which(AllDCNGeneb$chrom==x) ,c("sampleID", "start.pos", "cutoff")], start.pos~sampleID))
Xrs=lapply(Xl,function(x) rowSums(abs(x)))
Xind=unlist(lapply(Xrs, function(x) which(x>30)))

AllDCNGeneb2=AllDCNGeneb[which(AllDCNGeneb$start.pos%in%names(Xind)), ]
AllDCNGeneb2$mean[which(AllDCNGeneb2$mean>2.5)]=2.5
AllDCNGeneb2$mean[which(AllDCNGeneb2$mean<(-2.5))]=(-2.5)

```

## CIN calculation

CIN (Chromosomal instability) can be approximated by looking at the proportion of the genome/exome with copy number changes. We can calculate the degree of CIN by:

* Considering the fraction of sampled genome with called CN which is not 2
* Consider seg.mean FC greater than certain threshold (0.5)
* Consider seg.mean FC greater than certain threshold (0.5) multiplied by purity

Depending on the threshold used, the adjustment may completely invert the correlation. Using called CN results in less biased distribution

```{r cin-calc}
# using the same metric
Cin3a=sapply(CNVlog2, function(x) CalcGGI(x$C-2, 0.1, x$loc.end-x$loc.start))
Cin3b=sapply(CNVlog2, function(x) CalcGGI(x$C-2, 0.1))
Cin1a=sapply(CNVlog2, function(x) CalcGGI(x$seg.mean, 0.5))
Cin3d=sapply(1:length(GeneCNVs), function(x) CalcGGI(CNVlog2[[x]]$seg.mean, 0.5*pinfo[["Purity", x]]))

colIds=SampleIDs$OutCol[match(names(Cin1a), SampleIDs$SAMPLE_ID)]

par(mfrow=c(2,2))
plot(pinfo["Purity", ], Cin3a, xlab="purity", ylab="CIN (overall)", main="using CN",pch=19, col=as.character(colIds))
plot(pinfo["Purity", ], Cin3b, xlab="purity", ylab="CIN (overall)", main="using CN without gene length adjust",pch=19, col=as.character(colIds))
plot(pinfo["Purity", ], Cin1a, xlab="purity", ylab="CIN (overall)", main="T=0.5 before adjust",pch=19, col=as.character(colIds))
plot(pinfo["Purity", ], Cin3d, xlab="purity", ylab="CIN (overall)", main="T=0.5, purity  adjust",pch=19, col=as.character(colIds))
```

Using a segmented threshold of of 0.5 with purity adjustment shows a fairly even distribution.

## Cosmic Specific Genes

Below are CNV overviews of samples ranked by increasing CIN (approximated by the number of breakpoints observed). ER samples are highlighted, and data is truncated to lie within a [-2.5, 2.5] range

```{r, fig.cap="Overview of CNVs of hallmark genes" }
## melt everything and pull out the cosmic specific genes

#CNVmelt=melt(CNVlog2, measure.vars="end")
## order according to increasing CIN? as a metric: use number of variants: number of rows as an approximation
cin=sapply(CNVlog2, nrow)
cin2=sort(cin, decreasing = T)

AllDCNGene$sampleID=factor(AllDCNGene$sampleID, levels=names(cin2))
AllDCNGene$chrom=factor(AllDCNGene$chrom, levels=c(c(1:22)))
## pull out genes in the Cosmic List
getThese=which(AllDCNGene$gene%in%CmcHallmark$Gene.Symbol)
AllDCNGene2=AllDCNGene[getThese, ]
## sort the genes according to location
AllDCNGene2=AllDCNGene2[order(AllDCNGene2$chrom,AllDCNGene2$start.pos), ]
# set a hard threshold to visualise things easier
AllDCNGene2$mean[which(AllDCNGene2$mean>2.5)]=2.5
AllDCNGene2$mean[which(AllDCNGene2$mean<(-2.5))]=(-2.5)
AllDCNGene2$Alive_edited=S2$Alive_edited[match(AllDCNGene2$sampleID, S2$SAMPLE_ID)]
AllDCNGene2$cytoband=CmcHallmark$Chr.Band[match(AllDCNGene2$gene, CmcHallmark$Gene.Symbol)]

AllDCNGene2$sampleID=factor(AllDCNGene2$sampleID, levels=S2$SAMPLE_ID)
AllDCNGene2=AllDCNGene2[order(AllDCNGene2$sampleID), ]

AllDCNGene2$C2=AllDCNGene2$C
AllDCNGene2$C2[which(AllDCNGene2$C2>8)]=8
##
ggplot(AllDCNGene2, aes(y=sampleID, x=gene, fill=C2))+geom_tile()+facet_grid(Alive_edited~chrom, scale="free", space="free")+scale_fill_gradient2(low = "blue", mid="white", high = "red", na.value="white")+theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

##, axis.text.y = element_text(color =S2$ERautocol)
```

Most common gains are in the RTKs: KRAS, ERBB2, EGFR, MYC. Most common losses: SMADs, CDK2NA
We can load the Inputs from GISTIC to reveal which genes are significantly enriched:

```{r gistic-res}
AmpList=list()
getwd()
dir("data/gistic/all", "*90.txt")
AmpList[[1]]=read.delim(AllDat$Gistic$GoodResp_amp)
AmpList[[2]]=read.delim(AllDat$Gistic$PoorResp_amp)
AmpList[[3]]=read.delim(AllDat$Gistic$All_amp)

Mx2=unique(unlist(AmpList))

DelList=list()

DelList[[1]]=read.delim(AllDat$Gistic$GoodResp_del)
DelList[[2]]=read.delim(AllDat$Gistic$PoorResp_del)
DelList[[3]]=read.delim(AllDat$Gistic$All_del)

Mx3=unique(unlist(DelList))

AllDCNGene3=AllDCNGene2[which(AllDCNGene2$gene%in%c(Mx2, Mx3)), ]
##
AllDCNGene3$mean2=as.numeric(as.character(cut(AllDCNGene3$C, c(-1, 0.99, 1.01, 2.99, 3.99, 1000), c(-2, -1, 0, 1, 2))))

ggplot(AllDCNGene3, aes(y=sampleID, x=gene, fill=mean2))+geom_tile()+facet_grid(Alive_edited~chrom, scale="free", space="free")+scale_fill_gradient2(low = "blue", mid="white", high = "red", na.value="white")+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```

## Co-amplification patterns

Compute whether there are specific genes which are co-amplified or deleted? (Specifically assess the RTK genes for co-amp or deletions)

```{r}
TableCNVs=acast(AllDCNGene2[ ,c("sampleID", "gene", "C")], gene~sampleID)
NXTable=cor(t(TableCNVs))
heatmap.2(NXTable, scale="none", col=brewer.pal(11, "RdBu")[11:1], trace="none")

cor.test.p <- function(x){
    FUN <- function(x, y) cor.test(x, y)[["p.value"]]
    z <- outer(
      colnames(x), 
      colnames(x), 
      Vectorize(function(i,j) FUN(x[,i], x[,j]))
    )
    dimnames(z) <- list(colnames(x), colnames(x))
    z
}

corRes=cor.test.p(t(TableCNVs))
which(corRes<0.01)
```


## Summary of degree of CNVs across samples


Append a layer for CNV levels across the samples. In the 

```{r, eval=F}
ggplot(data=m2, aes(x=Sample, y=value,col=value))+geom_bar(stat="identity")+facet_grid(variable~Alive, space="free_x", scale="free")+theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))+ggtitle("Survival data: stage colored")+ylab("Number of variants")+theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))


## can merge with existing snv data:

mdfSumm2=rbind(mdfSumm[ , c("Sample","Alive_edited", "variable","value", "Alive")], m2[ , c("Sample","Alive_edited","variable","value", "Alive")])

head(mdfSumm2)

ggplot(data=mdfSumm2, aes(x=Sample, y=value,col=value))+geom_bar(stat="identity")+facet_grid(variable~Alive, space="free_x", scale="free")+theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))+ggtitle("Survival data: stage colored")+ylab("Number of variants")+theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

#pdf("~/Desktop/SNV_CV_summary.pdf", width=9, height=5)

ggplot(data=mdfSumm2, aes(x=Sample, y=value,col=value))+geom_bar(stat="identity")+facet_grid(variable~Alive_edited, space="free_x", scale="free")+theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))+ggtitle("Survival data: stage colored")+ylab("Number of variants")+theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

mdfSumm2$value[which(mdfSumm2$variable=="Somatic")]=log2(mdfSumm2$value[which(mdfSumm2$variable=="Somatic")])

ggplot(data=mdfSumm2, aes(x=Alive_edited, y=value, col=variable))+geom_boxplot()+facet_wrap(~variable, scales = "free")

t.test(mdfSumm2$value[which(mdfSumm2$Alive_edited=="Alive"&mdfSumm2$variable=="Somatic")], mdfSumm2$value[which(mdfSumm2$Alive_edited=="Deceased"&mdfSumm2$variable=="Somatic")])

t2=which(mdfSumm2$variable=="Germ" & mdfSumm2$value<5)
mdfSumm2[t2, ]
sample2=mdfSumm2$Sample[t2]
SampleIDs[match(sample2, SampleIDs$Name), ]

```
        